---
title: 'Statistical Computing: Project 3'
author: "Josh Stim"
date: "2023-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploring album sales and sentiment of lyrics from Beyoncé and Taylor Swift {.tabset .tabset-fade}

## Part 0. Setting up

Load relevant R packages
```{r}
library(here)
library(tidyverse)
```

Get data from tidytuesday github and store locally. Load the datasets and make copy for modification.
```{r}
if (!file.exists(here("data", "b_lyrics.RDS"))) {
    b_lyrics <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-29/beyonce_lyrics.csv")
    ts_lyrics <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-29/taylor_swift_lyrics.csv")
    sales <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-29/sales.csv")

    # save the files to RDS objects
    saveRDS(b_lyrics, file = here("data", "b_lyrics.RDS"))
    saveRDS(ts_lyrics, file = here("data", "ts_lyrics.RDS"))
    saveRDS(sales, file = here("data", "sales.RDS"))
}

b_lyrics.raw <- readRDS(here("data", "b_lyrics.RDS"))
ts_lyrics.raw <- readRDS(here("data", "ts_lyrics.RDS"))
sales.raw <- readRDS(here("data", "sales.RDS"))

b_lyrics.mod <- b_lyrics.raw
ts_lyrics.mod <- ts_lyrics.raw
sales.mod <- sales.raw
```

## Part 1. Explore album sales

In this section, the goal is to explore the sales of studio albums from Beyoncé and Taylor Swift.

#### Notes

-   In each of the subsections below that ask you to create a plot, you must create a title, subtitle, x-axis label, and y-axis label with units where applicable. For example, if your axis says “sales” as an axis label, change it to “sales (in millions)”.

### a. Data wrangling

1.    Use lubridate to create a column called released that is a Date class. However, to be able to do this, you first need to use stringr to search for pattern that matches things like this “(US)[51]” in a string like this “September 1, 2006 (US)[51]” and removes them. (Note: to get full credit, you must create the regular expression).

2.    Use forcats to create a factor called country (Note: you may need to collapse some factor levels).

3.    Transform the sales into a unit that is album sales in millions of dollars.

4.    Keep only album sales from the UK, the US or the World.

5.    Auto print your final wrangled tibble data frame.

I will start by taking a look at the sales dataset to see how many observations and variables we have, as well as the variable types. After taking a `glimpse()` of the data, I know we have ** 48 observations (or rows) ** and ** 8 variables (or columns) **. All variables are of type `<chr>`, except for `sales`, which is of type `<dbl>`.
```{r}
glimpse(sales.raw)
```

For the first part of ** 1a **, I need to convert the `released` column to a `date` class. Before doing this, I will need to make sure that all observations in the existing `released` column are of the same format.

We see here that this is not the case. That is, there are some observations that have an extra set of characters at the end (e.g., "September 1, 2006 (US)[51]").
```{r}
sales.raw$released
```

Here, I used `str_replace()` to modify the `released` variable in `sales.mod` and then converted the variable to type `Date`.
```{r}
sales.mod$released <- 
  str_replace(sales.raw$released, "( |)\\(U(K|S)\\)\\[[0-9]{2}\\]$", "") %>%
  mdy()

sales.mod$released
```

Now, we need to convert the `country` variable from `<chr>` to `<factor>`. We can begin this process by applying the `factor()` function and seeing what factors we end up with. For this project, I will assume that `FRA` and `FR` both refer to France, and `WW` and `World` both refer to "World Wide". 
```{r}
fct_count(sales.raw$country)
```

Here is my attempt to clean up the factors for `country`. As we can see, the `country` variable now has only 8 factors.
```{r}
sales.mod$country <- 
  factor(sales.raw$country) %>%
  fct_collapse(
    WW = c("WW", "World"),
    FRA = c("FRA","FR")
  )

fct_count(sales.mod$country)
```

Now, we will transform the `sales` variable from Dollars, to *Millions of Dollars*, and only include album sales from US, UK and WW.
```{r}
sales.mod$sales <- sales.raw$sales / (10 ** 6)
sales.mod.subset <- filter(sales.mod, country %in% c("US", "UK", "WW"))

sales.mod.subset
```
